graphsim: An R package for simulating gene expression data from graph structures of biological pathways

### 2020-05-25 paper/paper.Rmd
update code examples

```{r simulation_activating_setup, fig.align='center', fig.show='hold', fig.width='1.0\\linewidth', fig.height='1.0\\linewidth', out.height='.415\\linewidth', out.width='.415\\linewidth', fig.retina=10, fig.margin = FALSE, fig.ncol = 2, warning=FALSE}
# activating graph
state <- rep(1, length(E(graph)))
plot_directed(graph, state=state, layout = layout.kamada.kawai,
              cex.node=2, cex.arrow=4, arrow_clip = 0.2)
mtext(text = "(a) Activating pathway structure", side=1, line=3.5, at=0.075, adj=0.5, cex=1.75)
box()
#adjacency matrix
adj_mat <- make_adjmatrix_graph(graph)
#relationship matrix
dist_mat <- make_distance_graph(graph, absolute = FALSE)
#sigma matrix directly from graph
sigma_mat <- make_sigma_mat_dist_graph(graph, 0.8, absolute = FALSE)
#show shortest paths of graph
shortest_paths <- shortest.paths(graph)
#generate expression data directly from graph
expr <- generate_expression(100, graph, cor = 0.8, mean = 0, comm = FALSE,
                            dist = TRUE, absolute = FALSE, state = state)
```

```{r simulation_activating, fig.align='center', fig.show='hold', fig.width='1.0\\linewidth', fig.height='1.0\\linewidth', out.height='.415\\linewidth', out.width='.415\\linewidth', fig.retina=10, fig.margin = FALSE, fig.ncol = 2, fig.cap = '\\textbf{Simulating expression from a graph structure}. An example of a graph structure (a) that has been used to derive a relationship matrix (b), $\\Sigma$  matrix (c) and correlation structure (d) from the relative distances between the nodes. Non-negative values are coloured white to red from $0$ to $1$. This $\\Sigma$ matrix has been used to generate a simulated expression dataset of 100 samples (coloured blue to red from low to high) via sampling from the multivariate normal distribution. Here genes with closer relationships in the pathway structure show a higher correlation between simulated values.', warning=FALSE}

#plot relationship matrix
heatmap.2(make_distance_graph(graph, absolute = FALSE),
          scale = "none", trace = "none", col = colorpanel(50, "white", "red"),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)))
mtext(text = "(b) Relationship matrix", side=1, line=3.5, at=0, adj=0.5, cex=1.75)
#plot sigma matrix
heatmap.2(make_sigma_mat_dist_graph(graph, 0.8, absolute = FALSE),
scale = "none", trace = "none", col = colorpanel(50, "white", "red"),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)))
mtext(text = expression(paste("(c) ", Sigma, " matrix")), side=1, line=3.5, at=0, adj=0.5, cex=1.75)
#simulated data
expr <- generate_expression(100, graph, cor = 0.8, mean = 0,
comm = FALSE, dist =TRUE, absolute = FALSE, state = state)
#plot simulated correlations
heatmap.2(cor(t(expr)), scale = "none", trace = "none", col = colorpanel(50, "white", "red"),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)))
mtext(text = "(d) Simulated correlation", side=1, line=3.5, at=0, adj=0.5, cex=1.75)
#plot simulated expression data
heatmap.2(expr, scale = "none", trace = "none", col = bluered(50),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)), labCol = "")
mtext(text = "samples", side=1, line=1.5, at=0.2, adj=0.5, cex=1.5)
mtext(text = "genes", side=4, line=1, at=-0.4, adj=0.5, cex=1.5)
mtext(text = "(e) Simulated expression data (log scale)", side=1, line=3.5, at=0, adj=0.5, cex=1.75)
```

@@

```{r simulation_inhibiting_setup, fig.align='center', fig.show='hold', fig.width='1.0\\linewidth', fig.height='1.0\\linewidth', out.height='.415\\linewidth', out.width='.415\\linewidth', fig.retina=10, fig.margin = FALSE, fig.ncol = 2, warning=FALSE}
#generate parameters for inhibitions
state <- c(1, 1, -1, 1, 1, 1, 1, -1)
plot_directed(graph, state=state, layout = layout.kamada.kawai,
              cex.node=2, cex.arrow=4, arrow_clip = 0.2)
mtext(text = "(a) Inhibiting pathway structure", side=1, line=3.5, at=0.075, adj=0.5, cex=1.75)
box()
#adjacency matrix
adj_mat <- make_adjmatrix_graph(graph)
#relationship matrix
dist_mat <- make_distance_graph(graph, absolute = FALSE)
#sigma matrix directly from graph
sigma_mat <- make_sigma_mat_dist_graph(graph, state = state, 0.8, absolute = FALSE)
#show shortest paths of graph
shortest_paths <- shortest.paths(graph)
#generate expression data directly from graph
expr <- generate_expression(100, graph, state = state, cor = 0.8, mean = 0, comm = FALSE,
                            dist = TRUE, absolute = FALSE)
```
```{r simulation_inhibiting, fig.align='center', fig.show='hold', fig.width='1.0\\linewidth', fig.height='1.0\\linewidth', out.height='.415\\linewidth', out.width='.415\\linewidth', fig.retina=10, fig.margin = FALSE, fig.ncol = 2, fig.cap = '\\textbf{Simulating expression from graph structure with inhibitions}. An example of a graph structure (a), that has been used to derive a relationship matrix (b), $\\Sigma$ matrix (c), and correlation structure (d), from the relative distances between the nodes. These values are coloured blue to red from $-1$ to $1$. This has been used to generate a simulated expression dataset of 100 samples (coloured blue to red from low to high) via sampling from the multivariate normal distribution. Here the inhibitory relationships between genes are reflected in negatively correlated simulated  values.', warning=FALSE}
#plot relationship matrix
heatmap.2(make_distance_graph(graph, absolute = FALSE),
          scale = "none", trace = "none", col = colorpanel(50, "white", "red"),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)))
mtext(text = "(b) Relationship matrix", side=1, line=3.5, at=0, adj=0.5, cex=1.75)
# #plot sigma matrix
heatmap.2(make_sigma_mat_dist_graph(graph, state = state, 0.8, absolute = FALSE),
scale = "none", trace = "none", col = colorpanel(50, "blue", "white", "red"),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)))
mtext(text = expression(paste("(c) ", Sigma, " matrix")), side=1, line=3.5, at=0, adj=0.5, cex=1.75)
#simulated data
expr <- generate_expression(100, graph, state = state, cor = 0.8, mean = 0,
comm = FALSE, dist =TRUE, absolute = FALSE)
#plot simulated correlations
heatmap.2(cor(t(expr)), scale = "none", trace = "none", col = colorpanel(50, "blue", "white", "red"),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)))
mtext(text = "(d) Simulated correlation", side=1, line=3.5, at=0, adj=0.5, cex=1.75)
#plot simulated expression data
heatmap.2(expr, scale = "none", trace = "none", col = bluered(50),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)), labCol = "")
mtext(text = "samples", side=1, line=1.5, at=0.2, adj=0.5, cex=1.5)
mtext(text = "genes", side=4, line=1, at=-0.4, adj=0.5, cex=1.5)
mtext(text = "(e) Simulated expression data (log scale)", side=1, line=3.5, at=0, adj=0.5, cex=1.75)
```

@@

```{r simulation_smad_setup, fig.align='center', fig.show='hold', fig.width='1.0\\linewidth', fig.height='1.0\\linewidth', out.height='.415\\linewidth', out.width='.415\\linewidth', fig.retina=10, fig.margin = FALSE, fig.ncol = 2,  warning=FALSE}
#import graph from data
graph <- identity(TGFBeta_Smad_graph)
#generate parameters for inhibitions
state <- rep(1, length(E(graph)))
pathway <- get.edgelist(graph)
state[pathway[,1] %in% c("SMAD6", "SMAD7", "BAMBI", "SMURF1", "SMURF2", "UCHL5", "USP15", "UBB", "UBC", "PMEPA1", "PPP1CA", "PPP1CB", "PPP1CC", "PPP1R15A")] <- 2
state[is.na(state)] <- 1
plot_directed(graph, state = state, layout = layout.kamada.kawai,
              border.node=scales::alpha("black", 0.75), fill.node="lightblue",
              col.arrow = c(scales::alpha("navyblue", 0.25), scales::alpha("red", 0.25))[state], 
              cex.node = 1.5, cex.label = 0.8, cex.arrow = 2, 
              sub = expression(paste("(a) TFG-", beta, " activates SMADs")), cex.sub = 1.75)
box()
#adjacency matrix
adj_mat <- make_adjmatrix_graph(graph)
#relationship matrix
dist_mat <- make_distance_graph(graph, absolute = FALSE)
#sigma matrix directly from graph
sigma_mat <- make_sigma_mat_dist_graph(graph, state = state, 0.8, absolute = FALSE)
#show shortest paths of graph
shortest_paths <- shortest.paths(graph)
#generate expression data directly from graph
expr <- generate_expression(100, graph, state = state, cor = 0.8, mean = 0, comm = FALSE,
                            dist = TRUE, absolute = FALSE)
```
```{r simulation_smad, fig.align='center', fig.show='hold', fig.width='1.0\\linewidth', fig.height='1.0\\linewidth', out.height='.415\\linewidth', out.width='.415\\linewidth', fig.retina=10, fig.margin = FALSE, fig.ncol = 2, fig.cap = '\\textbf{Simulating expression from a biological pathway graph structure}. The graph structure (a) of a known biological pathway, the TGF-$\\beta$ receptor signaling activates SMADs (R-HSA-2173789), was used to derive a relationship matrix (b), $\\Sigma$ matrix (c) and correlation structure (d) from the relative distances between the nodes. These values are coloured blue to red from $-1$ to $1$. This has been used to generate a simulated expression dataset of 100 samples (coloured blue to red from low to high) via sampling from the multivariate normal distribution. Here modules of genes with correlated expression can be clearly discerned.', warning=FALSE}
# #plot relationship matrix
heatmap.2(make_distance_graph(graph, absolute = FALSE),
          scale = "none", trace = "none", col = colorpanel(50, "white", "red"),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)), labCol = "")
mtext(text = "(b) Relationship matrix", side=1, line=3.5, at=0, adj=0.5, cex=1.75)
# #plot sigma matrix
heatmap.2(make_sigma_mat_dist_graph(graph, state = state, 0.8, absolute = FALSE),
scale = "none", trace = "none", col = colorpanel(50, "blue", "white", "red"),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)), labCol = "")
mtext(text = expression(paste("(c) ", Sigma, " matrix")), side=1, line=3.5, at=0, adj=0.5, cex=1.75)
#simulated data
expr <- generate_expression(100, graph, state = state, cor = 0.8, mean = 0,
comm = FALSE, dist =TRUE, absolute = FALSE)
#plot simulated correlations
heatmap.2(cor(t(expr)), scale = "none", trace = "none", col = colorpanel(50, "blue", "white", "red"),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)), labCol = "")
mtext(text = "(d) Simulated correlation", side=1, line=3.5, at=0, adj=0.5, cex=1.75)
#plot simulated expression data
heatmap.2(expr, scale = "none", trace = "none", col = bluered(50),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)), labCol = "")
mtext(text = "samples", side=1, line=1.5, at=0.2, adj=0.5, cex=1.5)
mtext(text = "genes", side=4, line=1, at=-0.4, adj=0.5, cex=1.5)
mtext(text = "(e) Simulated expression data (log scale)", side=1, line=3.5, at=0, adj=0.5, cex=1.75)
```

>>>>

```{r simulation_activating, fig.align='center', fig.show='hold', fig.width='1.0\\linewidth', fig.height='1.0\\linewidth', out.height='.415\\linewidth', out.width='.415\\linewidth', fig.retina=10, fig.margin = FALSE, fig.ncol = 2, fig.cap = '\\textbf{Simulating expression from a graph structure}. An example of a graph structure (a) that has been used to derive a relationship matrix (b), $\\Sigma$  matrix (c) and correlation structure (d) from the relative distances between the nodes. Non-negative values are coloured white to red from $0$ to $1$. This $\\Sigma$ matrix has been used to generate a simulated expression dataset of 100 samples (coloured blue to red from low to high) via sampling from the multivariate normal distribution. Here genes with closer relationships in the pathway structure show a higher correlation between simulated values.', warning=FALSE}
# activating graph
state <- rep(1, length(E(graph)))
plot_directed(graph, state=state, layout = layout.kamada.kawai,
  cex.node=2, cex.arrow=4, arrow_clip = 0.2)
mtext(text = "(a) Activating pathway structure", side=1, line=3.5, at=0.075, adj=0.5, cex=1.75)
box()
#plot relationship matrix
heatmap.2(make_distance_graph(graph, absolute = FALSE),
  scale = "none", trace = "none", col = colorpanel(50, "white", "red"),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)))
mtext(text = "(b) Relationship matrix", side=1, line=3.5, at=0, adj=0.5, cex=1.75)
#plot sigma matrix
heatmap.2(make_sigma_mat_dist_graph(graph, 0.8, absolute = FALSE),
scale = "none", trace = "none", col = colorpanel(50, "white", "red"),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)))
mtext(text = expression(paste("(c) ", Sigma, " matrix")), side=1, line=3.5, at=0, adj=0.5, cex=1.75)
#simulated data
expr <- generate_expression(100, graph, cor = 0.8, mean = 0,
comm = FALSE, dist =TRUE, absolute = FALSE, state = state)
#plot simulated correlations
heatmap.2(cor(t(expr)), scale = "none", trace = "none", col = colorpanel(50, "white", "red"),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)))
mtext(text = "(d) Simulated correlation", side=1, line=3.5, at=0, adj=0.5, cex=1.75)
#plot simulated expression data
heatmap.2(expr, scale = "none", trace = "none", col = bluered(50),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)), labCol = "")
mtext(text = "samples", side=1, line=1.5, at=0.2, adj=0.5, cex=1.5)
mtext(text = "genes", side=4, line=1, at=-0.4, adj=0.5, cex=1.5)
mtext(text = "(e) Simulated expression data (log scale)", side=1, line=3.5, at=0, adj=0.5, cex=1.75)
```

@@

```{r simulation_inhibiting, fig.align='center', fig.show='hold', fig.width='1.0\\linewidth', fig.height='1.0\\linewidth', out.height='.415\\linewidth', out.width='.415\\linewidth', fig.retina=10, fig.margin = FALSE, fig.ncol = 2, fig.cap = '\\textbf{Simulating expression from graph structure with inhibitions}. An example of a graph structure (a), that has been used to derive a relationship matrix (b), $\\Sigma$ matrix (c), and correlation structure (d), from the relative distances between the nodes. These values are coloured blue to red from $-1$ to $1$. This has been used to generate a simulated expression dataset of 100 samples (coloured blue to red from low to high) via sampling from the multivariate normal distribution. Here the inhibitory relationships between genes are reflected in negatively correlated simulated  values.', warning=FALSE}
#generate parameters for inhibitions
state <- c(1, 1, -1, 1, 1, 1, 1, -1)
plot_directed(graph, state=state, layout = layout.kamada.kawai,
  cex.node=2, cex.arrow=4, arrow_clip = 0.2)
mtext(text = "(a) Inhibiting pathway structure", side=1, line=3.5, at=0.075, adj=0.5, cex=1.75)
box()
#plot relationship matrix
heatmap.2(make_distance_graph(graph, absolute = FALSE),
  scale = "none", trace = "none", col = colorpanel(50, "white", "red"),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)))
mtext(text = "(b) Relationship matrix", side=1, line=3.5, at=0, adj=0.5, cex=1.75)
# #plot sigma matrix
heatmap.2(make_sigma_mat_dist_graph(graph, state = state, 0.8, absolute = FALSE),
scale = "none", trace = "none", col = colorpanel(50, "blue", "white", "red"),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)))
mtext(text = expression(paste("(c) ", Sigma, " matrix")), side=1, line=3.5, at=0, adj=0.5, cex=1.75)
#simulated data
expr <- generate_expression(100, graph, state = state, cor = 0.8, mean = 0,
comm = FALSE, dist =TRUE, absolute = FALSE)
#plot simulated correlations
heatmap.2(cor(t(expr)), scale = "none", trace = "none", col = colorpanel(50, "blue", "white", "red"),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)))
mtext(text = "(d) Simulated correlation", side=1, line=3.5, at=0, adj=0.5, cex=1.75)
#plot simulated expression data
heatmap.2(expr, scale = "none", trace = "none", col = bluered(50),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)), labCol = "")
mtext(text = "samples", side=1, line=1.5, at=0.2, adj=0.5, cex=1.5)
mtext(text = "genes", side=4, line=1, at=-0.4, adj=0.5, cex=1.5)
mtext(text = "(e) Simulated expression data (log scale)", side=1, line=3.5, at=0, adj=0.5, cex=1.75)
```

@@

```{r simulation_smad, fig.align='center', fig.show='hold', fig.width='1.0\\linewidth', fig.height='1.0\\linewidth', out.height='.415\\linewidth', out.width='.415\\linewidth', fig.retina=10, fig.margin = FALSE, fig.ncol = 2, fig.cap = '\\textbf{Simulating expression from a biological pathway graph structure}. The graph structure (a) of a known biological pathway, the TGF-$\\beta$ receptor signaling activates SMADs (R-HSA-2173789), was used to derive a relationship matrix (b), $\\Sigma$ matrix (c) and correlation structure (d) from the relative distances between the nodes. These values are coloured blue to red from $-1$ to $1$. This has been used to generate a simulated expression dataset of 100 samples (coloured blue to red from low to high) via sampling from the multivariate normal distribution. Here modules of genes with correlated expression can be clearly discerned.', warning=FALSE}
#import graph from data
graph <- identity(TGFBeta_Smad_graph)
#generate parameters for inhibitions
state <- rep(1, length(E(graph))); pathway <- get.edgelist(graph)
state[pathway[,1] %in% c("SMAD6", "SMAD7", "BAMBI", "SMURF1", "SMURF2", "UCHL5",
  "USP15", "UBB", "UBC", "PMEPA1", "PPP1CA", "PPP1CB", "PPP1CC", "PPP1R15A")] <- 2
state[is.na(state)] <- 1
plot_directed(graph, state = state, layout = layout.kamada.kawai,
  border.node=scales::alpha("black", 0.75), fill.node="lightblue",
  col.arrow = c(scales::alpha("navyblue", 0.25), scales::alpha("red", 0.25))[state], 
  cex.node = 1.5, cex.label = 0.8, cex.arrow = 2, 
  sub = expression(paste("(a) TFG-", beta, " activates SMADs")), cex.sub = 1.75)
box()
# #plot relationship matrix
heatmap.2(make_distance_graph(graph, absolute = FALSE),
  scale = "none", trace = "none", col = colorpanel(50, "white", "red"),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)), labCol = "")
mtext(text = "(b) Relationship matrix", side=1, line=3.5, at=0, adj=0.5, cex=1.75)
# #plot sigma matrix
heatmap.2(make_sigma_mat_dist_graph(graph, state = state, 0.8, absolute = FALSE),
scale = "none", trace = "none", col = colorpanel(50, "blue", "white", "red"),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)), labCol = "")
mtext(text = expression(paste("(c) ", Sigma, " matrix")), side=1, line=3.5, at=0, adj=0.5, cex=1.75)
#simulated data
expr <- generate_expression(100, graph, state = state, cor = 0.8,
  mean = 0,comm = FALSE, dist =TRUE, absolute = FALSE)
#plot simulated correlations
heatmap.2(cor(t(expr)), scale = "none", trace = "none", col = colorpanel(50, "blue", "white", "red"),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)), labCol = "")
mtext(text = "(d) Simulated correlation", side=1, line=3.5, at=0, adj=0.5, cex=1.75)
#plot simulated expression data
heatmap.2(expr, scale = "none", trace = "none", col = bluered(50),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)), labCol = "")
mtext(text = "samples", side=1, line=1.5, at=0.2, adj=0.5, cex=1.5)
mtext(text = "genes", side=4, line=1, at=-0.4, adj=0.5, cex=1.5)
mtext(text = "(e) Simulated expression data (log scale)", side=1, line=3.5, at=0, adj=0.5, cex=1.75)
```
